# Extensions - Class options


--- 
subtitle: Penn Doctoral Weekend Presentation - 2022
site: bookdown::bookdown_site
---

### Overview of Gifted Options

|District-Wide Programs|City-Wide Programs|
|---|----|----|
|Qualifications (K-3)|Score 90 or higher (95+ for admission)|Score 97+ (99+ for admission)|
|Classroom format|Mixed (G&T + GenEd)|All classrooms for G&T|
|Boroughs w/ programs|All 5 boroughs|No school in Bronx & Staten Island|
|Student home location|Students living in the district have high priority|All parts of the city|


```{r message=FALSE, warning=FALSE, include=FALSE}
library(ggplot2)
library(ggmap)
```



### geoSpatial Mapping of Class Options

Each school's Economic Need Index (ENI) is the average of students' Economic Need Values. ENI measures estimates the percentages of students facing economic hardship.

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
X2019_2020_School_Point_Locations <- read_csv("raw/2019_-_2020_School_Point_Locations.csv")

### using the first column to identify school locations 
head(X2019_2020_School_Point_Locations)
```
```{r message=FALSE, warning=FALSE, include=FALSE}
#filter out schools with GT option
merged <- read_csv("/cloud/project/data/merged (1).csv")
head(merged)
nrow(merged)
merged_gf <- merged %>% filter(gifted == 1)
head(merged_gf)
```


```{r message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
merged_combined <- na.omit(X2019_2020_School_Point_Locations %>% 
                         left_join(merged, by = c("Loc_Name"= "School Name")) %>%
                         select(c("the_geom","Loc_Name","DBN","borough","Economic Need Index", "% Students with Disabilities")))
head(merged_combined)

gf_combined <- na.omit(X2019_2020_School_Point_Locations %>% 
                         left_join(merged_gf, by = c("Loc_Name"= "School Name")) %>%
                         select(c("the_geom","Loc_Name","DBN","borough","Economic Need Index")))

head(gf_combined)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
#change the_geom to separate columns of long and lat data
x <- str_extract(gf_combined$the_geom,"(?<=\\().*(?=\\))")
x <- unlist(strsplit(x," "))
```

```{r message=FALSE, warning=FALSE, include=FALSE}
#Extract long+lat data
odd <- seq(1,201,2)
even <- seq(2,202,2)
long <- c()
lat <- c()

for (i in 1:101){
  long[i] <- x[odd[i]]
  lat[i] <- x[even[i]]
}
gf_combined$long <- as.numeric(long)
gf_combined$lat <- as.numeric(lat)
gf_combined
```

```{r message=FALSE, warning=FALSE, include=FALSE}
#change the_geom to separate columns of long and lat data
y <- str_extract(merged_combined$the_geom,"(?<=\\().*(?=\\))")
y <- unlist(strsplit(y," "))

```

```{r message=FALSE, warning=FALSE, include=FALSE}
#Extract long+lat data
odd <- seq(1,2205,2)
even <- seq(2,2206,2)
long <- c()
lat <- c()

for (i in 1:1105){
  long[i] <- y[odd[i]]
  lat[i] <- y[even[i]]
}
merged_combined$long <- as.numeric(long)
merged_combined$lat <- as.numeric(lat)
merged_combined
```

```{r message=FALSE, warning=FALSE, include=FALSE}
merged_plot <- merged_combined %>% mutate(across(where(is.numeric), ~ round(., 3))) %>% select("Economic Need Index",long,lat)
names(merged_plot)[1] <- "ENI"
merged_plot$ENI <- merged_plot$ENI*100
merged_plot <- na.omit(merged_plot)
merged_plot
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
nyc.map <- get_map(location = "New York City", zoom = 10, maptype = "roadmap")
ggmap(nyc.map, extend = "device") +
  stat_summary_2d(data = merged_plot,aes(x = long, y = lat, z = ENI),alpha = 0.6)+
  scale_fill_gradient(name = "ENI", low = "green", high = "red")+
  geom_point(aes(x=long,y=lat,color=borough),data = gf_combined,size=0.7)
```


```{r message=FALSE, warning=FALSE, include=FALSE}
head(merged_combined)
merged_plot <- merged_combined %>% mutate(across(where(is.numeric), ~ round(., 3))) %>% select("% Students with Disabilities",long,lat)
names(merged_plot)[1] <- "Dis"
merged_plot$Dis <- merged_plot$Dis*100
merged_plot <- na.omit(merged_plot)
merged_plot
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
nyc.map2 <- get_map(location = "New York City", zoom = 10, maptype = "roadmap")
ggmap(nyc.map2, extend = "device") +
  stat_summary_2d(data = merged_plot,aes(x = long, y = lat, z = Dis),alpha = 0.6)+
  scale_fill_gradient(name = "% Dis", low = "green", high = "red")+
  geom_point(aes(x=long,y=lat,color=borough),data = gf_combined,size=0.7)
```



