# FE





```{r include=FALSE}

#install and load packages
# install.packages("tidyverse")
### load libraries
library(tidyverse)
library(readr)
library(tidyverse)
library(kableExtra)
library(kableExtra)
library(knitr)
library(broom)
library(viridis)

### bring in raw data
X2017_2018_SCHOOL_LEVEL_CLASS_SIZE_REPORT <- read_csv("/cloud/project/raw/2017-2018_SCHOOL_LEVEL_CLASS_SIZE_REPORT.csv")
#take out class size because that doesn't include D75
X2018_2019_School_Demographic_Snapshot <- read_csv("/cloud/project/raw/2018-2019_School_Demographic_Snapshot.csv")
X2013_2019_Attendance_Results_School <- read_csv("/cloud/project/raw/2013-2019_Attendance_Results_-_School.csv")

################
### Step I: Create the full_school data file
################
### step 1: from the demographics, get one row per DBN of just the demographics (they stay constant for 2017)
demos<- X2018_2019_School_Demographic_Snapshot %>%
  filter(Year == "2017-18") %>%
  select(DBN, `School Name`, `Total Enrollment`, `Economic Need Index`, '% Male', '% Black', '% White', '% Students with Disabilities', '% Poverty')
### step 2: merge the Grade + Program Type variables, then spread 
classes <- X2017_2018_SCHOOL_LEVEL_CLASS_SIZE_REPORT %>%
  mutate(program_type_grade = paste(`Program Type`, '-', `Grade Level`)) %>%
  select(DBN, program_type_grade, `Number of Classes`)%>%
  # mutate(grouped_id = row_number()) %>%
  spread(program_type_grade, `Number of Classes`)
#replace NA with 0
classes[is.na(classes)] <- 0
#rename columns
classes <- classes %>%
  rename(GT1 = "G&T - 1", GT2 = "G&T - 2", GT3 = "G&T - 3",
         GT4 = "G&T - 4", GT5 = "G&T - 5", GTK = "G&T - K",
         GE1 = "Gen Ed - 1",  GE2 = "Gen Ed - 2",  GE3 = "Gen Ed - 3", 
         GE4 = "Gen Ed - 4",  GE5 = "Gen Ed - 5",  GE6 = "Gen Ed - 6", 
         GE7 = "Gen Ed - 7",  GE8 = "Gen Ed - 8",  GEK = "Gen Ed - K", 
         ICT1 = "ICT - 1",   ICT2 = "ICT - 2",   ICT3 = "ICT - 3", 
         ICT4 = "ICT - 4",   ICT5 = "ICT - 5",   ICT6 = "ICT - 6",
         ICT7 = "ICT - 7",   ICT8 = "ICT - 8",ICTK = "ICT - K",
         ICTGT1 = "ICT & G&T - 1",  ICTGT2 = `ICT & G&T - 2`,  ICTGT3 = `ICT & G&T - 3`,
         ICTGT4 = "ICT & G&T - 4",  ICTGT5 = `ICT & G&T - 5`,  ICTGTK = `ICT & G&T - K`,
         SC121 = "SC 12:1 - K-8 SC", SC1211 = "SC 12:1:1 - K-8 SC", SC151 = "SC 15:1 - K-8 SC", 
         SC611 = "SC 6:1:1 - K-8 SC", SC811 = "SC 8:1:1 - K-8 SC")
### step 2A: create a binary indicator of whether middle school, K-8, or K-5
classes <- classes %>%
  mutate(middle_levels = ifelse((GE6 | GE7 | GE8 > 1) |(ICT6 | ICT7 | ICT8 > 1) , 1,0),
         elementary_levels = ifelse((GE1 | GE2 | GE3 |GE4| GE5 >1) | (ICT1 |ICT2|ICT3|ICT4|ICT5 > 1), 1,0),
         self_contained = ifelse(SC121 | SC1211 | SC151 | SC611 | SC811 > 1, 1,0),
         gifted = ifelse(GT1 | GT2 | GT3 | GT4 |GT5 |GTK >1, 1,0),
         self_contained_and_gifted = ifelse(self_contained ==1 & gifted ==1,1,0))
classes <- classes %>%
  mutate(school_type = case_when(middle_levels ==1 & elementary_levels ==0 ~ "Middle",
                                 middle_levels ==0 & elementary_levels ==1 ~ "Elementary",
                                 middle_levels == 1 & elementary_levels ==1 ~ "K to 8",
                                 middle_levels ==0 & elementary_levels ==0 ~ "gifted"),
         class_option = case_when(self_contained ==1 & gifted ==0 ~ "SC",
                                  self_contained ==1 & gifted ==1 ~ "SC and GT",
                                  self_contained ==0 & gifted == 0 ~ "no option",
                                  self_contained == 0 & gifted ==1 ~ "GT"))
classes <- classes %>%
  mutate(borough = case_when(str_detect(`DBN`, "X") ~"Bronx",
                             str_detect(`DBN`, "Q") ~ "Queens",
                             str_detect(`DBN`, "K") ~"Brooklyn",
                             str_detect(`DBN`, "M") ~ "Manhattan",
                             str_detect(`DBN`, "R") ~"Staten Island"))
## step 3: merge with the demographics
merged <- merge(demos, classes, by = "DBN") %>%
  distinct()

### step 4: get attendance
attendance<- X2013_2019_Attendance_Results_School %>%
  filter(`Grade` == "All Grades" & `Demographic Variable` == "All Students") %>%
  select(DBN, '% Attendance', '% Chronically Absent', 'Year') %>%
  filter(Year == '2017-18')
### merge with attendance
merged <- merge(merged, attendance, by = "DBN")
write.csv(merged, file = "/cloud/project/data/full_school_data.csv")
rm(list = ls())

```



### Data cleaning

Using OpenData NYC, we pull aggregate K-8 attendance records and socio-economic variables for all 1125 public schools.

```{r echo=FALSE, message=FALSE, warning=FALSE}
merged<- read_csv("data/full_school_data.csv")
head(merged[1:3,c(1:6,8:9)]) %>%
  kable(caption = "Sample Data Frame",
        col.names = c(" ", "DBN", "School Name",
                      "ENI", "% Male", "% Black", "%SWD", "% Poverty"))
```



```{r message=FALSE, warning=FALSE, include=FALSE}

library(tidyverse)
full_school_data <- read_csv("data/full_school_data.csv")
### edit the frame
working_data <- full_school_data %>%
  rename(TotalEnrollment = `Total Enrollment`,
         ENI = `Economic Need Index`,
         PercentBlack = `% Black`,
         PercentWhite = `% White`,
         PercentSWD = `% Students with Disabilities`,
         PercentPoverty = `% Poverty`,
         AllStudents_CA = `% Chronically Absent`,
         AllStudents_PA = `% Attendance`) %>%
  mutate(treatment = ifelse(self_contained ==0, 1,0))%>%
  select(DBN, `TotalEnrollment`, ENI, PercentBlack, PercentSWD, PercentPoverty, self_contained,
        AllStudents_CA, AllStudents_PA, treatment, borough, school_type,
        class_option) %>%
  na.omit()
## without matching, t.test is different in all of the covariates.
school_covariates <- c('ENI', 'PercentBlack', 'PercentSWD', 'PercentPoverty', 'TotalEnrollment')
with(working_data, t.test(AllStudents_PA ~ self_contained))
with(working_data, t.test(AllStudents_CA ~ self_contained))
with(working_data, t.test(ENI ~ self_contained))
with(working_data, t.test(PercentBlack ~ self_contained))
with(working_data, t.test(PercentSWD ~ self_contained))
with(working_data, t.test(PercentPoverty ~ self_contained))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
### t tests
## Attendance is sig different
working_data %>%
  group_by(treatment) %>%
  summarise(mean_attendance = mean(as.numeric(AllStudents_PA)),
            mean_chronic_absent = mean(as.numeric(AllStudents_CA))) %>%
  kable(caption = "Outcome Differences by Treatment",
        col.names= c("Treatment", "Mean Attendance", "Mean Chronic Absenteeism"),
        digits = c(0,2,2))
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
working_data %>%
  group_by(treatment) %>%
  select(one_of(school_covariates)) %>%
  summarise_all(funs(mean(., na.rm=T)))%>%
  kable(caption = "Covariate Differences by Treatment",
        col.names= c("Treatment", "ENI", "Percent Black",
                     "PercentSWD", "Percent Poverty", "Total Enrollment"),
        digits = c(0,2,3,2,3, 3))

```



### Covariates

Treatment (inclusion) is a significant predictor of absenteeism (CA), without matching. However, five covariates generally associated with higher absenteeism are predictive of both the treatment and the outcome, making them confounders.

$$
\begin{align}
ChronicAbsenteeism = \beta_0 +\beta_1Treatment + \beta_2ENI+ 
\beta_3PercentBlack+\beta_4PercentSWD + \\ \beta_5PercentPoverty +\beta_6PercentPoverty+ BoroughFixedEffects + \epsilon
\end{align}
$$




```{r echo=FALSE}

## multiple regression

lm(AllStudents_CA ~ treatment + 
              TotalEnrollment + ENI + PercentBlack +PercentSWD +PercentPoverty +
              factor(borough), data = working_data) %>%
  tidy() %>%
  kable(caption =  "Chronic Absenteeism ~ Treatment + Covariates",
        col.names = c("Predictor", "B", "SE", "t", "p"),
        digits = c(0,2,3,2,3))
```


$$
\begin{align}
Treatment = \beta_0 +\beta_1TotalEnrollment + \beta_2ENI+ 
\beta_3PercentBlack+\beta_4PercentSWD + \\ \beta_5PercentPoverty + BoroughFixedEffects + \epsilon
\end{align}
$$

```{r echo=FALSE}
## Predict treatment
lm(treatment ~  TotalEnrollment + ENI + PercentBlack +PercentSWD +PercentPoverty + factor(borough), data = working_data) %>%
  tidy() %>%
  kable(caption =  "Treatment ~ Covariates",
        col.names = c("Predictor", "B", "SE", "t", "p"),
        digits = c(0,2,3,2,3))



```


## Matching

We match at a ratio of 3:1, control:treatment groups. The original sample has 885 Control (no inclusion) schools and 240 Treatment (inclusion) schools. The matched sample has 418 Control and 181 Treatment. All covariates cross the acceptable standardized mean difference threshold, and visual diagnostics of the covariate distributions confirm the balance improvement.

### Assess Balance

```{r echo=FALSE, fig.show="hold", fig.width=5, fig.height = 3, message=FALSE, warning=FALSE}
merged<- read_csv("data/full_school_data.csv")
#libraries
library(MatchIt)
### nearest matching
merged$treatment <- ifelse(merged$self_contained == 1, 0 , 1)
merged <- merged %>%
  rename(TotalEnrollment = `Total Enrollment`,
         ENI = `Economic Need Index`,
         PercentBlack = `% Black`,
         PercentWhite = `% White`,
         PercentSWD = `% Students with Disabilities`,
         PercentPoverty = `% Poverty`) %>%
  select(treatment, `TotalEnrollment`, ENI, PercentBlack, 
         PercentSWD, PercentPoverty, borough, `% Chronically Absent`, `% Attendance`)
  
school_nearest <- matchit(treatment ~ PercentBlack + PercentSWD + PercentPoverty + TotalEnrollment +
                          ENI, 
                          family = binomial(),
                          data = merged,
                          method = "nearest",
                           caliper = 0.25,
                           ratio = 3)

#create the matched set
nearest_matched <- match.data(school_nearest)
library(cobalt)
bal.tab(school_nearest, m.threshold = 0.1)
bal.plot(school_nearest, var.name = 'TotalEnrollment', which = "both")
bal.plot(school_nearest, var.name = 'PercentPoverty', which = "both")
bal.plot(school_nearest, var.name = 'PercentBlack', which = "both")
bal.plot(school_nearest, var.name = 'PercentSWD', which = "both")
bal.plot(school_nearest, var.name = 'ENI', which = "both")

```

### Matched Estimates

Matching yields a statistically significant treatment effect for both absenteeism and attendance, improving our confidence in the causal effect of inclusion on overall school attendance rates.


$$
\begin{align}
ChronicAbsenteeism = \beta_0 +\beta_1Treatment + \epsilon
\end{align}
$$

```{r echo=FALSE, message=FALSE, warning=FALSE}
## estimating treatment effects
lm(`% Chronically Absent` ~ treatment + PercentBlack + PercentSWD +
     borough, data = nearest_matched) %>%
  tidy() %>%
  kable(caption =  "Causal Effect using Matched Data",
        col.names = c("Predictor", "B", "SE", "t", "p"),
        digits = c(0,2,3,2,3))

```


### within borough matching


```{r echo=FALSE, fig.show="hold", fig.width=5, fig.height = 3, message=FALSE, warning=FALSE}

bronx <- merged %>%
  filter(borough == "Bronx")
school_nearest_bronx <- matchit(treatment ~ PercentSWD + PercentPoverty + TotalEnrollment +
                          ENI, 
                          family = binomial(),
                          data = bronx,
                          method = "nearest",
                           caliper = 0.25,
                           ratio = 2)

manhattan <- merged %>%
  filter(borough == "Manhattan")
school_nearest_manhattan <- matchit(treatment ~  PercentBlack+ PercentSWD + PercentPoverty + TotalEnrollment +
                          ENI, 
                          family = binomial(),
                          data = manhattan,
                          method = "nearest",
                           caliper = 0.25,
                           ratio = 2)


# Bronx
school_nearest <- school_nearest_bronx
nearest_matched <- match.data(school_nearest)
library(cobalt)
bal.tab(school_nearest, m.threshold = 0.1)
## estimating treatment effects
lm(`% Chronically Absent` ~ treatment, data = nearest_matched) %>%
  tidy() %>%
  kable(caption =  "Causal Effect using Matched Data",
        col.names = c("Predictor", "B", "SE", "t", "p"),
        digits = c(0,2,3,2,3))



# Manhattan
school_nearest <- school_nearest_manhattan
nearest_matched <- match.data(school_nearest)
library(cobalt)
bal.tab(school_nearest, m.threshold = 0.1)
## estimating treatment effects
lm(`% Chronically Absent` ~ treatment, data = nearest_matched) %>%
  tidy() %>%
  kable(caption =  "Causal Effect using Matched Data",
        col.names = c("Predictor", "B", "SE", "t", "p"),
        digits = c(0,2,3,2,3))

```




#### match on boroughs

Two ways:multilevel logistic regression and logistic regression with fixed cluster effects can be used to estimate propensity scores. The effects of individual-level covariates vary across clusters. 


```{r}
merged<- read_csv("data/full_school_data.csv")
#libraries
library(MatchIt)
### nearest matching
merged$treatment <- ifelse(merged$self_contained == 1, 0 , 1)
merged <- merged %>%
  rename(TotalEnrollment = `Total Enrollment`,
         ENI = `Economic Need Index`,
         PercentBlack = `% Black`,
         PercentWhite = `% White`,
         PercentSWD = `% Students with Disabilities`,
         PercentPoverty = `% Poverty`) %>%
  select(treatment, `TotalEnrollment`, ENI, PercentBlack, 
         PercentSWD, PercentPoverty, borough, `% Chronically Absent`, `% Attendance`)
  
school_nearest <- matchit(treatment ~  PercentBlack+ PercentSWD + PercentPoverty + TotalEnrollment +
                          ENI + borough, 
                          family = binomial(),
                          data = merged,
                          method = "nearest",
                           caliper = 0.25,
                           ratio = 2)

library(cobalt)
bal.tab(school_nearest, m.threshold = 0.1)
nearest_matched <- match.data(school_nearest)
## estimating treatment effects
lm(`% Chronically Absent` ~ treatment +  borough, data = nearest_matched) %>%
  tidy() %>%
  kable(caption =  "Causal Effect using Matched Data",
        col.names = c("Predictor", "B", "SE", "t", "p"),
        digits = c(0,2,3,2,3))
```




